---
title: "Time normalization for IntelliCages"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Roland Bock"
date: "`r format(Sys.time(), '%A %B %d, %Y')`"
output: 
  html_document:
    fig_caption: yes
    fig_width: 10
    highlight: tango
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Problem

The [IntelliCages][1] by [TSE-systems][2] provide an automated system of group-housed mouse behavior over long periods of time, only interrupted by required cage changes. If one of the desired analysis is the drinking behavior over time, it is desirable to express the time in relation to the light-dark cycle. In many cases, this can be simply achieved by relabeling the time axis according to the light-dark cycle. In other cases, as in the one that prompted this solution, it is not quite as straight forward. In this example, the recording went through the time shift from standard time to daylight savings time, i.e. the light-dark cycle within the facility did not change, but the computer automatically (per default) adjusted it's own time to the daylight savings time. Although nothing changed for the individual mouse, a histogram on the visit or nosepoke times will show a time shift. 

## Prerequisite

This is not an explanation of the full functionality of the Intellicages. The example data for this solution are exported text files from the IntelliCage Analyzer Program, which creates 7 or 8 distinct files, but only a subset are included here: 

-  **Animal**.txt      - summary for each animal
-  **Environment**.txt - enviroment data for temperature and illumination
-  **Visit**.txt      - summary of the visit events

For detailed explanations, please refer to the help sections on your own IntelliCage system.

## Goal

The goal of these steps is to recalculate each visit timestamp relative to the beginning of its dark cycle. The same procedure can be used with the nosepokes, which are not included here. 

## Problem visualized

To start I am going to visualize the problem. Let's begin with loading some neccessary packages.

```{r packages, message=FALSE, warning=FALSE}
library(easypackages)
packages("tidyverse", "patchwork", "kableExtra", "here", "lubridate",
         "labelled", "hms")
```

### Animals

Next load the animal data, so that we can use the assigned groups for the visit data. This data is loaded from a separate csv file to make reassigning the groups easier.

```{r load-data}

subjects <- read_csv(file = here("data", "animal_overview.csv"), col_names = T,
                       col_type = "ccccic")
subjects$Group <- factor(subjects$Group, levels = c("control", "autokn", "idkn", "ddkn"))
subjects$DOB <- as.Date(subjects$DOB, "%m/%d/%y")
subjects <- rename(subjects, Tag = RFID)
var_label(subjects) <- list(Animal = "Mouse ID number",
                              Sex = "Sex",
                              DOB = "Date of birth",
                              Tag = "Chip ID number",
                              Cohort = "Cohort number",
                              Group = "Experimental group")

subjects %>% 
  arrange(Group, Sex) %>% 
  kable(caption = "Overview of experimental animals") %>% 
  kable_classic(lightable_options = "hover", font_size = 14, full_width = F)
```

### Visits

Loading the _Visit.txt_ file with the summary data for each registered visit.

```{r load-visits, message=FALSE, warning=FALSE}
visits.col.spec <- cols(
  .default = col_integer(),
  VisitID = col_integer(),
  VisitOrder = col_integer(),
  Animal = col_character(),
  Tag = col_character(),
  Sex = col_character(),
  Group = col_character(),
  Module = col_character(),
  Cage = col_integer(),
  Corner = col_integer(),
  CornerCondition = col_factor(), 
  PlaceError = col_logical(),
  SideErrors = col_double(),
  TimeErrors = col_double(),
  ConditionErrors = col_double(), 
  NosepokeNumber = col_double(),
  NosepokeDuration = col_double(),
  LickNumber = col_double(),
  LickDuration = col_double(),
  LickContactTime = col_double(),
  StartDate = col_date(),
  StartTime = col_time(),
  StartTimecode = col_double(),
  EndDate = col_date(),
  EndTime = col_time(),
  EndTimecode = col_double(),
  VisitDuration = col_double(),
  Session = col_integer()
)

visitData <- read_tsv(file = here("data/Cohort 1/", "Visit.txt"), col_types = visits.col.spec)

tibble(variables = ncol(visitData),
       sessions = nlevels(as.factor(visitData$Session)),
       start = min(visitData$StartDate),
       end = max(visitData$EndDate),
       visits = format(nrow(visitData), big.mark = ",", justify = "right"),
       pokes = format(sum(visitData$NosepokeNumber), big.mark = ","),
       licks = format(sum(visitData$LickNumber), big.mark = ",")) %>%
  kable(caption = "Quick visit data overview") %>%
  kable_classic(font_size = 16)

```

In the next step I am combining the _subject_ data frame with the _visitData_ data frame, mainly to update the grouping. 

```{r g-alcohol-histogram, fig.align="center", fig.cap="Alcohol drinking pattern of 4 groups of mice during the day with a reversed light-dark cycle over the course of 6 weeks. Darkgray bar indicates recorded dark cycle during standard time, lightgray bar shows dark cycle during daylight savings time. The light cycle did not change for the mice, the computer adjusted its recording time by default."}
visitsAll <- subjects %>% 
  select(Tag, Group) %>%            # select only the Tag and Group column from the subjects
  left_join(visitData %>% select(-c(Group)), by = "Tag") %>%      # join with visit data and drop the group column
  mutate(drink = if_else(Session > 1 & (Corner == 2 | Corner == 4), "alcohol", "water")) # introduce a drink 
                                                                                         # label for the corners
visitsAll %>% 
  filter(LickNumber > 0 & drink == "alcohol") %>% 

  ggplot(aes(x = StartTime, weight = LickNumber)) +
  geom_histogram(bins = 96) +
  geom_rect(aes(xmin = as_hms("6:30:00"), 
                xmax = as_hms("18:30:00"), ymin = -80, ymax = -10), 
            color = "white", fill = "lightgray") +
  geom_rect(aes(xmin = as_hms("5:30:00"), 
                xmax = as_hms("17:30:00"), ymin = -160, ymax = -90), 
            color = "white", fill = "darkgray") +
  scale_x_time(name = "time of day (h)", 
               breaks = seq(0, 24 * 3600, 6 * 3600), 
               labels = seq(0, 24, 6)) +
  theme_minimal() +
  theme(strip.background = element_rect(color = "lightgray", fill = "lightgray"),
        strip.text = element_text(face = "italic"),
        panel.spacing = unit(2, "lines")) +
  facet_grid(. ~ Group)
```

Ideally the recording becomes independent from the computer time and can be expressed as a relative unit from the beginning of the dark or light cycle.

## Solution

Luckily the IntelliCages have an build-in illumination and temperature sensor. The default setting records the illumination and temperature once a minute, but that can be changed by the user. Using the functionality of the [tidyverse][3], we can extract the start times of the dark cycle, merge them back into the visit or nosepoke data file and calculate the individual offset for each time.

The data for the temperature and illumination are stored in the _environment.txt_ file. It can be loaded as follows:

```{r load-env-data, warning=FALSE}
environment.col.spec <- cols(
  .default = col_integer(),
  DateTime = col_datetime(),
  Temperature = col_double(),
  Illumination = col_integer(),
  Cage = col_integer()
)

env <- read_tsv(file = here("data/Cohort 1/", "Environment.txt"),
                         col_types = environment.col.spec) %>% 
  mutate(date = ymd(format(DateTime, "%Y-%m-%d")),
         time = as_hms(format(DateTime, "%H:%M:%S")))

glimpse(env)
```

The recorded data should be quite the same for both cages, since they are in the same room. I am also just interested in the illumination value, so we can further drop the _Temperature_ and the _DateTime_ column.

```{r reduce-env-data}
env <- env %>% filter(Cage == 1) %>% select(Illumination, date, time)
```

This reduces the tibble to `r nrow(env)` rows and `r ncol(env)` columns.

### Workflow

#### Find all start times of the dark cycle

```{r find-start-dark}
envDark <- env %>% 
  mutate(Illumination = if_else(Illumination < 5, 0, 10),      # clean Illumination values and
         IllChange = Illumination - lag(Illumination))  %>%    #    calculate the change
  filter(IllChange < 0) %>% 
  mutate(darkStart = time)                                     # need an extra column for start time

envDark %>% 
  slice_head(n = 5) %>% 
  kable(caption = "First rows of new environmental data table.") %>% 
  kable_paper(font_size = 14, full_width = F)
```

As a good measure, lets perform quick sanity check to make sure all data points are consistent.

```{r g-starttimes, fig.align="center", fig.width=3, fig.height=4, fig.cap="A single outlier of the start times indicate that somebody turned the light on during the dark cycle."}
envDark %>% 
  ggplot(aes(x = 1, y = darkStart)) +
  geom_jitter(width = 0.2, shape = 19, fill = "gray", alpha = 0.5) +
  scale_x_continuous(name = "", limits = c(0, 2)) +
  scale_y_time(name = "start times of dark cycle",
               limits = c(5 * 3600, 11.2 * 3600)) +
  theme_minimal()
```

Let's remove the single outlier before we merge the start times back into the visits data frame.

```{r remove-outlier}
envDark <- envDark %>% filter(time < as_hms("10:00:00"))
```


I am interested in the start of the dark cycle, though the intial start of the program is in the middle of the dark cyle. We need to find the first start of the light cycle, subtract 12 hours from it and add it to the top of the table.

```{r corr-dark-times}
env %>% 
  mutate(Illumination = if_else(Illumination < 5, 0, 10),      # clean Illumination values and
         IllChange = Illumination - lag(Illumination))  %>%    #    calculate the change
  filter(IllChange > 0) %>%               # find all the light starts
  slice_head() %>%                        # limit to the first one
  mutate(time = as_hms(time - 12*60*60),  # subtract 12 hours from start time
         IllChange = -IllChange,
         Illumination = 0,
         darkStart = time) %>% 
  bind_rows(., envDark) -> envDark

envDark %>% 
  slice_head(n = 5) %>% 
  kable(caption = "Added data table with first dark cycle") %>% 
  kable_paper(font_size = 14, full_width = F)
```

#### Add dark cycle times to timestamps for visits

Now we need to get the time stamps for the visit data and match them with the correct dark cycle start time.

```{r create-visit-timestamps}
visitsDT <- visitsAll %>% 
  select(StartDate, StartTime) %>% 
  rename(date = "StartDate", time = "StartTime")

visitsDT <- visitsDT %>% 
  bind_rows(envDark) %>% 
  arrange(date, time) %>%                  # sort all data rows by date and time 
  fill(darkStart, .direction = "down") %>% # fill the missing values
  select(-c(Illumination, IllChange))      # columns are not needed anymore
```

#### Use the dark cycle start times to adjust start times

```{r corrected-visit-times}
visitsCor <- visitsAll %>% 
  inner_join(visitsDT, by = c(StartDate = "date", StartTime = "time")) %>% 
  mutate(dark = as_hms(StartTime - darkStart))
```

### Corrected Graph

```{r g-corrected-graph, fig.align="center", fig.cap="Alcohol drinking data plotted against corrected times. Time 0 is the start of the dark cycle."}
visitsCor %>% 
  filter(LickNumber > 0 & drink == "alcohol") %>% 

  ggplot(aes(x = dark, weight = LickNumber)) +
  geom_histogram(bins = 96) +
  geom_rect(aes(xmin = as_hms("0:00:00"), 
                xmax = as_hms("12:00:00"), ymin = -80, ymax = -10), 
            color = "white", fill = "darkgray") +
  scale_x_time(name = "time (h)", 
               breaks = seq(-6 * 3600, 18 * 3600, 6 * 3600), 
               labels = seq(-6, 18, 6)) +
  theme_minimal() +
  theme(strip.background = element_rect(color = "lightgray", fill = "lightgray"),
        strip.text = element_text(face = "italic"),
        panel.spacing = unit(2, "lines")) +
  facet_grid(. ~ Group)

```

As some might notice, the adjustment produces negative times for the ranges between midnight and the start of the dark cycle (assuming a constant 12 hour light on-light off cycle). If this is an unwanted effect, we can simply add the negative times to 24 hours (and subtract one day from the date) to align them to the 24 hour cycle.

```{r g-full-cycle, fig.align="center", fig.cap="Histogram with times aligned to the 24 hour cycle."}
visitsCor %>% 
  mutate(dark_d = if_else(dark < 0, StartDate - 1, StartDate),
         dark_t = if_else(dark < 0, as_hms(24 * 3600 + dark), dark)) %>% 
  filter(LickNumber > 0 & drink == "alcohol") %>% 

  ggplot(aes(x = dark_t, weight = LickNumber)) +
  geom_histogram(bins = 96) +
  geom_rect(aes(xmin = as_hms("0:00:00"), 
                xmax = as_hms("12:00:00"), ymin = -80, ymax = -10), 
            color = "white", fill = "darkgray") +
  scale_x_time(name = "time (h)", 
               breaks = seq(0, 24 * 3600, 6 * 3600), 
               labels = seq(0, 24, 6)) +
  theme_minimal() +
  theme(strip.background = element_rect(color = "lightgray", fill = "lightgray"),
        strip.text = element_text(face = "italic"),
        panel.spacing = unit(2, "lines")) +
  facet_grid(. ~ Group)

```


[1]: https://www.tse-systems.com/product-details/intellicage/
[2]: https://www.tse-systems.com/
[3]: https://www.tidyverse.org/